---
title: 'Allen: covariates and batch effects'
author: "Davide Risso"
date: "2/10/2016"
output: 
  html_document: 
    fig_height: 7
    fig_width: 7
    toc: yes
    code_folding: hide
    toc_float: yes
---

```{r options, echo=FALSE, results="hide",mesasge=FALSE, error=FALSE, include=FALSE, autodep=TRUE}
knitr::opts_chunk$set(fig.align="center", cache=TRUE, error=FALSE, message=FALSE, warning=TRUE)
library(SummarizedExperiment)
library(scRNAseq)
library(zinbwave)
library(ggplot2)
library(magrittr)
library(matrixStats)
library(Rtsne)
library(scone)
library(RColorBrewer)
library(cluster)
library(parallel)
library(EDASeq)
library(digest)
library(edgeR)
library(DESeq2)

set.seed(8283)
options(getClass.msg=FALSE) #get rid of annoying messages about cache until fixed internally in R

ncores <- detectCores() - 1
```

Here, I will explore the relation between the expression values and some covariates provided with the data.

```{r datain}
data("allen")
allen_core <- allen[grep("^ERCC-", rownames(allen), invert = TRUE),
                    which(colData(allen)$Core.Type=="Core" &
                            !(colData(allen)$Primary.Type %in% c("Pvalb Tacr3", "Sst Myh8")))]

filter <- rowSums(assay(allen_core)>10)>=10
```

# Comparison of methods

```{r normalization}
raw <- assay(allen_core)[filter,]

totalcount = function (ei)
{
  sums = colSums(ei)
  eo = t(t(ei)*mean(sums)/sums)
  return(eo)
}

tc <- totalcount(raw)
fq <- FQT_FN(raw)
tmm <- TMM_FN(raw)
deseq <- DESEQ_FN(raw)
```

## PCA

```{r pca}
library(rARPACK)
fastpca <- function(expr, scale=FALSE) {
  svd_raw <- svds(scale(t(expr), center=TRUE, scale=scale), k=3, nu=3, nv=0)
  pc_raw <- svd_raw$u %*% diag(svd_raw$d[1:3])
  return(pc_raw)
}

vars <- rowVars(log1p(tc))
names(vars) <- rownames(tc)
vars <- sort(vars, decreasing = TRUE)
vargenes <- names(vars)[1:1000]

pc_raw <- fastpca(log1p(raw[vargenes,]))
pc_tc <- fastpca(log1p(tc[vargenes,]))
pc_fq <- fastpca(log1p(fq[vargenes,]))
pc_tmm <- fastpca(log1p(tmm[vargenes,]))
pc_deseq <- fastpca(log1p(deseq[vargenes,]))

col1 <- brewer.pal(9, "Set1")
col2 <- c(brewer.pal(8, "Set2"), brewer.pal(8, "Set3"), brewer.pal(8, "Set1"))

level1 <- as.factor(colData(allen_core)$driver_1_s)
level2 <- as.factor(colData(allen_core)$Primary.Type)

colMerged <- col1[level1]
colCl <- col2[level2]

plot(pc_raw, col=colMerged, pch=20, main="PCA RAW")
legend("bottomright", levels(level1), fill=col1)

plot(pc_raw, col=colCl, pch=20, main="PCA RAW")
legend("bottomright", levels(level2), fill=col2)

plot(pc_tc, col=colMerged, pch=20, main="PCA TC")
plot(pc_tc, col=colCl, pch=20, main="PCA TC")
plot(pc_tmm, col=colMerged, pch=20, main="PCA TMM")
plot(pc_tmm, col=colCl, pch=20, main="PCA TMM")
plot(pc_fq, col=colMerged, pch=20, main="PCA FQ")
plot(pc_fq, col=colCl, pch=20, main="PCA FQ")
```

## ZIFA

```{r zifa}
wrapRzifa <- function(Y, block = T){
  # wrapper R function for ZIFA.
  # md5 hashing and temporary files are used not to re-run zifa 
  # if it has already be run on this computer.
  d = digest(Y, "md5")
  tmp = paste0(tempdir(), '/', d)
  write.csv(Y, paste0(tmp, '.csv'))
  
  if (!file.exists(paste0(tmp, '_zifa.csv'))){
    print('run ZIFA')
    bb = ifelse(block, '-b ', '')
    cmd = sprintf('python run_zifa.py %s%s.csv %s_zifa.csv', bb, tmp, tmp)
    system(cmd)
  }
  read.csv(sprintf("%s_zifa.csv", tmp), header=FALSE)
}

zifa_raw <- wrapRzifa(log1p(raw[vargenes,]))
plot(zifa_raw, pch=19, col=colMerged, main="ZIFA Raw")
plot(zifa_raw, pch=19, col=colCl, main="ZIFA Raw")

zifa_tc <- wrapRzifa(log1p(tc[vargenes,]))
plot(zifa_tc, pch=19, col=colMerged, main="ZIFA TC")
plot(zifa_tc, pch=19, col=colCl, main="ZIFA TC")

zifa_tmm <- wrapRzifa(log1p(tmm[vargenes,]))
plot(zifa_tmm, pch=19, col=colMerged, main="ZIFA TMM")
plot(zifa_tmm, pch=19, col=colCl, main="ZIFA TMM")

zifa_fq <- wrapRzifa(log1p(fq[vargenes,]))
plot(zifa_fq, pch=19, col=colMerged, main="ZIFA FQ")
plot(zifa_fq, pch=19, col=colCl, main="ZIFA FQ")
```

## ZINB

```{r}
library(doParallel)
library(BiocParallel)
NCORES = 2
registerDoParallel(NCORES)
register(DoparParam())
```

```{r zinb}
zinb <- zinbFit(raw[vargenes,], K = 2)
plot(zinb@W, pch=19, col=colMerged, main="ZINB")
plot(zinb@W, pch=19, col=colCl, main="ZINB")
```

# Sample-level covariates

## PCA

Here, we look at PCA (using FQ normalization) to see which covariates influence the data.

```{r pca_cov}
detection_rate <- colSums(raw>0)
coverage <- colSums(raw)

df <- cbind(data.frame(PC1=pc_fq[,1], PC2=pc_fq[,2], coverage=coverage, detection_rate=detection_rate), as.data.frame(colData(allen_core)[,1:15]))

df %>% ggplot(aes(PC1, PC2)) + geom_point(aes(color=coverage)) + scale_colour_gradient(low="blue", high="yellow") + theme_classic()

df %>% ggplot(aes(PC1, PC2)) + geom_point(aes(color=detection_rate)) + scale_colour_gradient(low="blue", high="yellow") + theme_classic()

pairs(df[,1:4], col=colMerged, pch=19)

cors <- sapply(1:3, function(i) cor(pc_tc[,i], df[,3:19]))
colnames(cors) <- paste("PC", 1:NCOL(cors), sep="")
rownames(cors) <- paste("QC", 1:NROW(cors), sep="")
barplot(abs(cors), beside = TRUE, col=rainbow(17), border=rainbow(17), ylim=c(0, 1), space=c(0, 2), main="Absolute correlation with QC features")
legend("topright", colnames(df)[3:19], fill=rainbow(17), border=rainbow(17), cex=.5)
```

## ZIFA

```{r zifa_cov}
df <- cbind(data.frame(Z1=zifa_fq[,1], Z2=zifa_fq[,2], coverage=coverage, detection_rate=detection_rate), as.data.frame(colData(allen_core)[,1:15]))

df %>% ggplot(aes(Z1, Z2)) + geom_point(aes(color=coverage)) + scale_colour_gradient(low="blue", high="yellow") + theme_classic()

df %>% ggplot(aes(Z1, Z2)) + geom_point(aes(color=detection_rate)) + scale_colour_gradient(low="blue", high="yellow") + theme_classic()

pairs(df[,1:4], col=colMerged, pch=19)

cors <- sapply(1:2, function(i) cor(zifa_tc[,i], df[,3:19]))
colnames(cors) <- paste("PC", 1:NCOL(cors), sep="")
rownames(cors) <- paste("QC", 1:NROW(cors), sep="")
barplot(abs(cors), beside = TRUE, col=rainbow(17), border=rainbow(17), ylim=c(0, 1), space=c(0, 2), main="Absolute correlation with QC features")
legend("topright", colnames(df)[3:19], fill=rainbow(17), border=rainbow(17), cex=.5)
```

## ZINB

```{r zinb_cov}
df <- cbind(data.frame(W1=zinb@W[,1], W2=zinb@W[,2], coverage=coverage, detection_rate=detection_rate), as.data.frame(colData(allen_core)[,1:15]))

df %>% ggplot(aes(W1, W2)) + geom_point(aes(color=coverage)) + scale_colour_gradient(low="blue", high="yellow") + theme_classic()

df %>% ggplot(aes(W1, W2)) + geom_point(aes(color=detection_rate)) + scale_colour_gradient(low="blue", high="yellow") + theme_classic()

pairs(df[,1:4], col=colMerged, pch=19)

cors <- sapply(1:2, function(i) cor(zinb@W[,i], df[,3:19]))
colnames(cors) <- paste("PC", 1:NCOL(cors), sep="")
rownames(cors) <- paste("QC", 1:NROW(cors), sep="")
barplot(abs(cors), beside = TRUE, col=rainbow(17), border=rainbow(17), ylim=c(0, 1), space=c(0, 2), main="Absolute correlation with QC features")
legend("topright", colnames(df)[3:19], fill=rainbow(17), border=rainbow(17), cex=.5)
```

## Add QC features in ZINB model

```{r zinb_batch}
qcpca <- prcomp(as.matrix(colData(allen_core)[,1:15]), center=TRUE, scale=TRUE)
zinb_batch <- zinbFit(raw[vargenes,], K = 2, X=qcpca$x[,1:2])
plot(zinb_batch@W, pch=19, col=colMerged, main="ZINB")
```

```{r zinb_batch_cov}
plot(zinb_batch@W, col=colMerged, pch=20, main="Colored by batch")

df <- cbind(data.frame(W1=zinb_batch@W[,1], W2=zinb_batch@W[,2], coverage=coverage, detection_rate=detection_rate), as.data.frame(colData(allen_core)[,1:15]))

df %>% ggplot(aes(W1, W2)) + geom_point(aes(color=coverage)) + scale_colour_gradient(low="blue", high="yellow") + theme_classic()

df %>% ggplot(aes(W1, W2)) + geom_point(aes(color=detection_rate)) + scale_colour_gradient(low="blue", high="yellow") + theme_classic()

pairs(df[,1:4], col=colMerged, pch=19)

cors <- sapply(1:2, function(i) cor(zinb_batch@W[,i], df[,3:19]))
colnames(cors) <- paste("PC", 1:NCOL(cors), sep="")
rownames(cors) <- paste("QC", 1:NROW(cors), sep="")
barplot(abs(cors), beside = TRUE, col=rainbow(17), border=rainbow(17), ylim=c(0, 1), space=c(0, 2), main="Absolute correlation with QC features")
legend("topright", colnames(df)[3:19], fill=rainbow(17), border=rainbow(17), cex=.5)
```

## Add layers in ZINB model

```{r zinb_bio}
cluster2 <- as.character(level2)
cluster2[cluster2 %in% c("Pvalb Tacr3", "Sst Myh8")] <- "Other"
cluster2[grep("^L4", cluster2)] <- "L4"
cluster2[grep("^L6", cluster2)] <- "L6a"
cluster2[grep("^L5a", cluster2)] <- "L5a"
cluster2[cluster2 == "L5 Ucma"] <- "L5a"
cluster2[grep("^L5b", cluster2)] <- "L5b"
cluster2 <- as.factor(cluster2)

zinb_bio <- zinbFit(raw[vargenes,], K = 2, X=model.matrix(~cluster2), epsilon=1e3)
plot(zinb_bio@W, pch=19, col=colMerged, main="ZINB")
```

```{r zinb_bio_cov, dependson="zinb_bio"}
plot(zinb_bio@W, col=col2[level2], pch=20, main="Colored by layer")
legend("bottomright", levels(level2), fill=col2, cex=.5)

df <- cbind(data.frame(W1=zinb_bio@W[,1], W2=zinb_bio@W[,2], coverage=coverage, detection_rate=detection_rate), as.data.frame(colData(allen_core)[,1:15]))

df %>% ggplot(aes(W1, W2)) + geom_point(aes(color=coverage)) + scale_colour_gradient(low="blue", high="yellow") + theme_classic()

df %>% ggplot(aes(W1, W2)) + geom_point(aes(color=detection_rate)) + scale_colour_gradient(low="blue", high="yellow") + theme_classic()

pairs(df[,1:4], col=colMerged, pch=19)

cors <- sapply(1:2, function(i) cor(zinb_bio@W[,i], df[,3:19]))
colnames(cors) <- paste("PC", 1:NCOL(cors), sep="")
rownames(cors) <- paste("QC", 1:NROW(cors), sep="")
barplot(abs(cors), beside = TRUE, col=rainbow(17), border=rainbow(17), ylim=c(0, 1), space=c(0, 2), main="Absolute correlation with QC features")
legend("topright", colnames(df)[3:19], fill=rainbow(17), border=rainbow(17), cex=.5)
```

```{r zinb_outliers, dependson="zinb_bio"}
outliers <- which(zinb_bio@W[,1]>2)
plot(zinb_bio@W, col="grey", pch=19)
points(zinb_bio@W[outliers,], col="red", pch=19)

qc <- as.matrix(colData(allen_core)[,1:15])

col_out <- rep("grey", nrow(qc))
col_out[outliers] <- "red"

plot(qc[,c(1, 4)], col=col_out, pch=19)

library(cluster)
pamres <- pam(zinb_bio@W[-outliers,], k=2)
table(pamres$clustering)

plot(zinb_bio@W[-outliers,], col=col1[pamres$clustering], pch=19)
```

```{r zinb_check_batch,eval=FALSE}
metadata <- read.csv("../../scone_analysis/data/allen_metadata.csv", stringsAsFactors = FALSE)
ids <- read.table("../../scone_analysis/data/allen_ids.txt", stringsAsFactors = FALSE, header = TRUE)

ids <- ids[match(colnames(raw), ids[,1]),]
batch_info <- metadata[,c("collection_date", "short_name", "notes")]
batch <- factor(batch_info[match(ids[,2], batch_info[,2]),1],
                levels=c("7/8/2013", "11/22/2013", "11/25/2013", "12/12/2013",
                         "1/9/2014", "1/31/2014", "2/10/2014", "3/27/2014",
                         "4/23/2014", "8/18/2014", "11/20/2014",
                         "11/25/2014", "2/11/2015", "2/18/2015"))
notes <- as.factor(batch_info[match(ids[,2], batch_info[,2]),3])
table(batch[-outliers], pamres$clustering)

plot(zinb_bio@W[-outliers,], col=col1[batch[-outliers]], pch=19)

boxplot(zinb_bio@W[,1]~batch, las=2)
boxplot(zinb_bio@W[-outliers,1]~batch[-outliers], las=2)
```

## Add batches in ZINB model

```{r zinb_batch2, dependson="zinb_check_batch",eval=FALSE}
zinb_batch <- zinbFit(raw[vargenes,], K = 2, X=model.matrix(~batch), epsilon=1e4)
plot(zinb_batch@W, pch=19, col=colMerged, main="ZINB")
plot(zinb_batch@W, pch=19, col=colCl, main="ZINB")
```

```{r zinb_batch2_cov, dependson="zinb_batch2",eval=FALSE}
plot(zinb@W, col=col1[batch], pch=19, main="Colored by batch")

df <- cbind(data.frame(W1=zinb_batch@W[,1], W2=zinb_batch@W[,2], coverage=coverage, detection_rate=detection_rate), as.data.frame(colData(allen_core)[,1:15]))

df %>% ggplot(aes(W1, W2)) + geom_point(aes(color=coverage)) + scale_colour_gradient(low="blue", high="yellow") + theme_classic()

df %>% ggplot(aes(W1, W2)) + geom_point(aes(color=detection_rate)) + scale_colour_gradient(low="blue", high="yellow") + theme_classic()

pairs(df[,1:4], col=colMerged, pch=19)

cors <- sapply(1:2, function(i) cor(zinb_batch@W[,i], df[,3:19]))
colnames(cors) <- paste("PC", 1:NCOL(cors), sep="")
rownames(cors) <- paste("QC", 1:NROW(cors), sep="")
barplot(abs(cors), beside = TRUE, col=rainbow(17), border=rainbow(17), ylim=c(0, 1), space=c(0, 2), main="Absolute correlation with QC features")
legend("topright", colnames(df)[3:19], fill=rainbow(17), border=rainbow(17), cex=.5)

```

# Gene-level covariates

```{r gc_content, eval=FALSE}
counts <- counts[bm[,1],]
gene_info = getGeneLengthAndGCContent(as.character(bm[,2]), "mm10", mode="org.db")
rownames(gene_info) = bm[,1]
gene_info = na.omit(gene_info)
```

```{r write_out, dependson=c("zinb", "zifa", "pca", "zinb_batch2", "zinb_bio"), eval=FALSE}
save(zinb, zinb_batch, zinb_bio, zifa_fq, zifa_tmm, zifa_tc, zifa_raw, pc_deseq, pc_tmm, pc_fq, pc_tc, pc_raw, 
file="allen_covariates_1000.rda")
```

```{r write_out_bis, dependson=c("zinb", "zifa", "pca", "zinb_bio")}
save(zinb, zinb_bio, zifa_fq, zifa_tmm, zifa_tc, zifa_raw, pc_deseq, pc_tmm, pc_fq, pc_tc, pc_raw, 
file="allen_covariates_1000.rda")
```
