---
title: "68k PBMC cells from 10x genomics"
author: "Davide Risso"
date: "07/07/2017"
output: 
  html_document: 
    fig_height: 7
    fig_width: 7
    toc: yes
    code_folding: hide
    toc_float: yes
---

```{r options, echo=FALSE, results="hide",mesasge=FALSE, error=FALSE, include=FALSE, autodep=TRUE}
knitr::opts_chunk$set(fig.align="center", cache=TRUE, error=FALSE, message=FALSE, warning=TRUE)
library(zinbwave)
library(ggplot2)
library(magrittr)
library(matrixStats)
library(Rtsne)
library(RColorBrewer)
library(cowplot)

set.seed(8283)
ncores <- 7
```

Here, we apply ZINB-WaVE to the set of 68k PBMCs from 10x genomics. The idea is
to show scalability of the method and hopefully new biological insights.

In the original pubblications, the authors performed k-means on the first 50 PCs
and identified 10 clusters. Here, I copy/paste snippets of text from their results.

- Clusters 1-3, 6: T cells (enriched in CD3D)
- Cluster 5: NK cells (NKG7)
- Cluster 7: B cells (CD79A)
- Cluster 9: myeloid cells (S100A8 and S100A9)
- Cluster 1, 4, 6: CD8+ cytotoxic T cells
- Cluster 2, 3 CD4+ T cells

Moreover, cells in cluster 3 showed high expression of CCR10 and TNFRSF18, markers for memory T cells and regulatory T cells respectively, and likely consisted of a mixture of memory and regulatory T cells. The presence of ID3, which is important in maintaining a naive T-cell state, suggests that cluster 2 represents naive CD8 T cells, whereas cluster 4 represents naive CD4 T cells.

To identify subpopulations within the myeloid population, they applied k-means clustering on the first 50 PCs of cluster 9 cells. At least three populations were evident: dendritic cells (characterized by the presence of FCER1A25), CD16þ monocytes and CD16-/low monocytes.

Cluster 8 showed preferential expression of megakaryocyte markers, such as PF4, suggesting that it represents a cluster of megakaryocytes. Cells in cluster 10 express markers of B, T and dendritic cells, suggesting a likely cluster of multiplets.

We classified 68k PBMCs based on their best match to the average expression profile of 11 reference transcriptomes. Cell classification was largely consistent with previously described marker-based classification, although the boundaries among some of the T-cell subpopulations were blurred. Namely, part of the inferred CD4+ naive T population was classified as CD8+ T cells. We have also tried to cluster the 68k PBMC data with Seurat. While it was able to distinguish inferred CD4+ naive from inferred CD8+ naive T cells, it was not able to cleanly separate out inferred activated cytotoxic T cells from inferred NK cells.

Interestingly, there is no mention of normalization in the paper. Were the data normalized?

Here, I download the original UMI count matrix and run ZINB-WaVE on it. We can consider two strategies: run zinbwave with K=2 or 3 for visualization or run it with an higher K for clustering (perhaps followed by t-SNE?).

I will visualize the data using the original clustering.

# Original analysis

I've downloaded the t-sne coordinates and cluster labels of the original analysis.
This should allow us to reproduce the original Figure 3.

```{r}
tsne_coords <- read.csv("tenx_pbmc/projection.csv", row.names=1)
clusters <- read.csv("tenx_pbmc/clusters.csv")
cl <- clusters[,2]
names(cl) <- clusters[,1]
table(cl)
dim(clusters)
dim(tsne_coords)

stopifnot(all(names(cl)==rownames(tsne_coords)))

tsne_coords$cluster <- as.factor(cl)
```

For some reason the downloaded data include only 40k cells instead of the advertised 68k...

```{r}
pal <- clusterExperiment::bigPalette
plot(tsne_coords, pch=19, col=pal[cl])

ggplot(aes(TSNE.1, TSNE.2, colour=cluster), data = tsne_coords) + geom_point()
```

To see if the clusters correspond to the published clusters, we need to look at
some marker genes. To do that, I need to load the full data.

```{r}
library(cellrangerRkit)
pipestance_path <- "tenx_pbmc"
if(!file.exists(paste0(pipestance_path, "/outs"))) {
  download_sample(sample_name="fresh_68k_pbmc_donor_a",sample_dir=pipestance_path,
                  host="http://cf.10xgenomics.com/samples/cell-exp/1.1.0/")
}
gbm <- load_cellranger_matrix(pipestance_path)
dim(gbm)
```

This object has the full dataset. For now, let's use the 40k cells for which they
provide the full analysis.

```{r}
all(rownames(tsne_coords) %in% colnames(gbm))
gbm_40k <- gbm[, rownames(tsne_coords)]
dim(gbm_40k)

tsne_coords$CD3D <- exprs(gbm_40k)[which(fData(gbm_40k)[,2] == "CD3D"),]
tsne_coords$NKG7 <- exprs(gbm_40k)[which(fData(gbm_40k)[,2] == "NKG7"),]
tsne_coords$CD79A <- exprs(gbm_40k)[which(fData(gbm_40k)[,2] == "CD79A"),]
tsne_coords$S100A8 <- exprs(gbm_40k)[which(fData(gbm_40k)[,2] == "S100A8"),]
tsne_coords$S100A9 <- exprs(gbm_40k)[which(fData(gbm_40k)[,2] == "S100A9"),]


ggplot(aes(TSNE.1, TSNE.2, colour=CD3D), data = tsne_coords) + 
  geom_point() + scale_colour_gradient(low="blue", high="yellow")
boxplot(tsne_coords$CD3D~cl)

ggplot(aes(TSNE.1, TSNE.2, colour=NKG7), data = tsne_coords) + 
  geom_point() + scale_colour_gradient(low="blue", high="yellow")
boxplot(tsne_coords$NKG7~cl)

ggplot(aes(TSNE.1, TSNE.2, colour=CD79A), data = tsne_coords) + 
  geom_point() + scale_colour_gradient(low="blue", high="yellow")
boxplot(tsne_coords$CD79A~cl)

ggplot(aes(TSNE.1, TSNE.2, colour=S100A8), data = tsne_coords) + 
  geom_point() + scale_colour_gradient(low="blue", high="yellow")
boxplot(tsne_coords$S100A8~cl)

ggplot(aes(TSNE.1, TSNE.2, colour=S100A9), data = tsne_coords) + 
  geom_point() + scale_colour_gradient(low="blue", high="yellow")
boxplot(tsne_coords$S100A9~cl)

```

There is no correspondance between the publication clusters and the clusters downloaded 
here.

# ZINB-WaVE

Here, I try to run zinbwave with K=2 on the full dataset. Following the authors'
analysis and our own practice, I start with the 1000 most variable genes.

```{r}
use_genes <- get_nonzero_genes(gbm)
dense <- as.matrix(exprs(gbm[use_genes,]))
vars <- rowVars(log1p(dense))
names(vars) <- rownames(dense)
vars <- sort(vars, decreasing = TRUE)
vargenes <- names(vars)[1:1000]

dense <- dense[vargenes,]
```

To avoid memory problems, I'm running it in serial mode for now.

```{r}
library(BiocParallel)
register(SerialParam())
system.time(zinb_res <- zinbFit(dense, K=2))
```
