---
title: 'Zeisel: covariates and batch effects'
author: "Davide Risso"
date: "11/17/2016"
output: 
  html_document: 
    fig_height: 7
    fig_width: 7
    toc: yes
    code_folding: hide
    toc_float: yes
---

```{r options, echo=FALSE, results="hide",mesasge=FALSE, error=FALSE, include=FALSE, autodep=TRUE}
knitr::opts_chunk$set(fig.align="center", cache=TRUE, error=FALSE, message=FALSE, warning=TRUE)
library(SummarizedExperiment)
library(scRNAseq)
library(zinb)
library(ggplot2)
library(magrittr)
library(matrixStats)
library(Rtsne)
library(scone)
library(RColorBrewer)
library(cluster)
library(parallel)
library(EDASeq)
library(digest)
library(edgeR)
library(DESeq2)

set.seed(8283)
options(getClass.msg=FALSE) #get rid of annoying messages about cache until fixed internally in R

ncores <- 23
```

Here, I will explore the relation between the expression values and some covariates provided with the data.
I will use both RPKM and full-quantile normalized data.

```{r datain}
data <- read.table("expression_mRNA_17-Aug-2014.txt", sep='\t', stringsAsFactors = FALSE, comment.char = '%')

tissue <- as.factor(as.matrix(data)[1,-(1:2)])
table(tissue)
group <- as.factor(as.matrix(data)[2,-(1:2)])
table(tissue, group)
nmolecule <- as.numeric(as.matrix(data)[3,-(1:2)])
well <- as.factor(as.matrix(data)[4,-(1:2)])
sex <- as.factor(as.matrix(data)[5,-(1:2)])
age <- as.numeric(as.matrix(data)[6,-(1:2)])
table(tissue, sex)
table(tissue, age)
diameter <- as.numeric(as.matrix(data)[7,-(1:2)])
cell_id <- as.matrix(data)[8,-(1:2)]
batch <- as.factor(sapply(strsplit(cell_id, "_"), function(x) x[1]))
position <- as.factor(sapply(strsplit(cell_id, "_"), function(x) x[2]))

level1 <- as.factor(as.matrix(data)[9,-(1:2)])
level2 <- as.factor(as.matrix(data)[10,-(1:2)])
table(level1)
table(level1, level2)

table(level1, group)

counts <- as.matrix(data[12:NROW(data),-(1:2)])
counts <- matrix(as.numeric(counts), ncol=ncol(counts), nrow=nrow(counts))
rownames(counts) <- data[12:NROW(data),1]
colnames(counts) <- data[8, -(1:2)]
```

# Comparison of methods

```{r normalization}
filter <- rowSums(counts>10)>=10
raw <- counts[filter,]

totalcount = function (ei)
{
  sums = colSums(ei)
  eo = t(t(ei)*mean(sums)/sums)
  return(eo)
}

tc <- totalcount(raw)
fq <- FQT_FN(raw)
tmm <- TMM_FN(raw)
deseq <- DESEQ_FN(raw)
```

## PCA

```{r pca}
library(rARPACK)
fastpca <- function(expr, scale=FALSE) {
  svd_raw <- svds(scale(t(expr), center=TRUE, scale=scale), k=3, nu=3, nv=0)
  pc_raw <- svd_raw$u %*% diag(svd_raw$d[1:3])
  return(pc_raw)
}

pc_raw <- fastpca(log1p(raw))
pc_tc <- fastpca(log1p(tc))
pc_fq <- fastpca(log1p(fq))
pc_tmm <- fastpca(log1p(tmm))
#pc_deseq <- fastpca(log1p(deseq))

col1 <- brewer.pal(9, "Set1")
col2 <- brewer.pal(8, "Set2")

colMerged <- col2[level1]
plot(pc_raw, col=colMerged, pch=20, main="PCA RAW")
legend("topleft", levels(level1), fill=col2)

plot(pc_tc, col=colMerged, pch=20, main="PCA TC")
plot(pc_tmm, col=colMerged, pch=20, main="PCA TMM")
plot(pc_fq, col=colMerged, pch=20, main="PCA FQ")
```

## ZIFA

```{r zifa}
wrapRzifa <- function(Y, block = T){
  # wrapper R function for ZIFA.
  # md5 hashing and temporary files are used not to re-run zifa 
  # if it has already be run on this computer.
  d = digest(Y, "md5")
  tmp = paste0(tempdir(), '/', d)
  write.csv(Y, paste0(tmp, '.csv'))
  
  if (!file.exists(paste0(tmp, '_zifa.csv'))){
    print('run ZIFA')
    bb = ifelse(block, '-b ', '')
    cmd = sprintf('python run_zifa.py %s%s.csv %s_zifa.csv', bb, tmp, tmp)
    system(cmd)
  }
  read.csv(sprintf("%s_zifa.csv", tmp), header=FALSE)
}

zifa_raw <- wrapRzifa(log1p(raw))
plot(zifa_raw, pch=19, col=colMerged, main="ZIFA Raw")

zifa_tc <- wrapRzifa(log1p(tc))
plot(zifa_tc, pch=19, col=colMerged, main="ZIFA TC")

zifa_tmm <- wrapRzifa(log1p(tmm))
plot(zifa_tmm, pch=19, col=colMerged, main="ZIFA TMM")

zifa_fq <- wrapRzifa(log1p(fq))
plot(zifa_fq, pch=19, col=colMerged, main="ZIFA FQ")
```

## ZINB

```{r zinb}
zinb <- zinbFit(raw, ncores = ncores, K = 2)
plot(zinb@W, pch=19, col=colMerged, main="ZINB")
```

# Sample-level covariates

## PCA

Here, we look at PCA (using TC normalization) to see which covariates influence the data.

```{r pca_cov}
detection_rate <- colSums(raw>0)
coverage <- colSums(raw)

colPal1 <- colorRampPalette(c("blue", "yellow"))(nlevels(batch)) 
colBatch <- colPal1[batch]
plot(pc_tc, col=colBatch, pch=20, main="Colored by batch")

colPal2 <- colorRampPalette(c("blue", "yellow"))(nlevels(position)) 
colPos <- colPal2[position]
plot(pc_tc, col=colPos, pch=20, main="Colored by position")

df <- data.frame(PC1=pc_tc[,1], PC2=pc_tc[,2], nmolecule=nmolecule, diameter=diameter, coverage=coverage, detection_rate=detection_rate) 

df %>% ggplot(aes(PC1, PC2)) + geom_point(aes(color=nmolecule)) + scale_colour_gradient(low="blue", high="yellow") + theme_classic()

df %>% ggplot(aes(PC1, PC2)) + geom_point(aes(color=diameter)) + scale_colour_gradient(low="blue", high="yellow") + theme_classic()

df %>% ggplot(aes(PC1, PC2)) + geom_point(aes(color=coverage)) + scale_colour_gradient(low="blue", high="yellow") + theme_classic()

df %>% ggplot(aes(PC1, PC2)) + geom_point(aes(color=detection_rate)) + scale_colour_gradient(low="blue", high="yellow") + theme_classic()

pairs(df, col=colMerged, pch=19)

cors <- sapply(1:3, function(i) cor(pc_tc[,i], df[,3:6]))
colnames(cors) <- paste("PC", 1:NCOL(cors), sep="")
rownames(cors) <- paste("QC", 1:NROW(cors), sep="")
barplot(abs(cors), beside = TRUE, col=col1[1:4], border=col1[1:4], ylim=c(0, 1), space=c(0, 2), main="Absolute correlation with QC features")
legend("topright", colnames(df)[3:6], fill=col1, border=col1, cex=1)

boxplot(pc_tc[,1]~batch, col=colPal1, ylab="PC1", xlab="Batch")
boxplot(pc_tc[,2]~batch, col=colPal1, ylab="PC2", xlab="Batch")

boxplot(pc_tc[,1]~position, col=colPal2, ylab="PC1", xlab="Position")
boxplot(pc_tc[,2]~position, col=colPal2, ylab="PC2", xlab="Position")
```

## ZIFA

```{r zifa_cov}
plot(zifa_tc, col=colBatch, pch=20, main="Colored by batch")

plot(zifa_tc, col=colPos, pch=20, main="Colored by position")

df <- data.frame(Z1=zifa_tc[,1], Z2=zifa_tc[,2], nmolecule=nmolecule, diameter=diameter, coverage=coverage, detection_rate=detection_rate) 

df %>% ggplot(aes(Z1, Z2)) + geom_point(aes(color=nmolecule)) + scale_colour_gradient(low="blue", high="yellow") + theme_classic()

df %>% ggplot(aes(Z1, Z2)) + geom_point(aes(color=diameter)) + scale_colour_gradient(low="blue", high="yellow") + theme_classic()

df %>% ggplot(aes(Z1, Z2)) + geom_point(aes(color=coverage)) + scale_colour_gradient(low="blue", high="yellow") + theme_classic()

df %>% ggplot(aes(Z1, Z2)) + geom_point(aes(color=detection_rate)) + scale_colour_gradient(low="blue", high="yellow") + theme_classic()

pairs(df, col=colMerged, pch=19)

cors <- sapply(1:2, function(i) cor(zifa_tc[,i], df[,3:6]))
colnames(cors) <- paste("PC", 1:NCOL(cors), sep="")
rownames(cors) <- paste("QC", 1:NROW(cors), sep="")
barplot(abs(cors), beside = TRUE, col=col1[1:4], border=col1[1:4], ylim=c(0, 1), space=c(0, 2), main="Absolute correlation with QC features")
legend("topright", colnames(df)[3:6], fill=col1, border=col1, cex=1)

boxplot(zifa_tc[,1]~batch, col=colPal1, ylab="PC1", xlab="Batch")
boxplot(zifa_tc[,2]~batch, col=colPal1, ylab="PC2", xlab="Batch")

boxplot(zifa_tc[,1]~position, col=colPal2, ylab="PC1", xlab="Position")
boxplot(zifa_tc[,2]~position, col=colPal2, ylab="PC2", xlab="Position")
```

## ZINB

```{r zinb_cov}
plot(zinb@W, col=colBatch, pch=20, main="Colored by batch")

plot(zinb@W, col=colPos, pch=20, main="Colored by position")

df <- data.frame(W1=zinb@W[,1], W2=zinb@W[,2], nmolecule=nmolecule, diameter=diameter, coverage=coverage, detection_rate=detection_rate) 

df %>% ggplot(aes(W1, W2)) + geom_point(aes(color=nmolecule)) + scale_colour_gradient(low="blue", high="yellow") + theme_classic()

df %>% ggplot(aes(W1, W2)) + geom_point(aes(color=diameter)) + scale_colour_gradient(low="blue", high="yellow") + theme_classic()

df %>% ggplot(aes(W1, W2)) + geom_point(aes(color=coverage)) + scale_colour_gradient(low="blue", high="yellow") + theme_classic()

df %>% ggplot(aes(W1, W2)) + geom_point(aes(color=detection_rate)) + scale_colour_gradient(low="blue", high="yellow") + theme_classic()

pairs(df, col=colMerged, pch=19)

cors <- sapply(1:2, function(i) cor(zinb@W[,i], df[,3:6]))
colnames(cors) <- paste("PC", 1:NCOL(cors), sep="")
rownames(cors) <- paste("QC", 1:NROW(cors), sep="")
barplot(abs(cors), beside = TRUE, col=col1[1:4], border=col1[1:4], ylim=c(0, 1), space=c(0, 2), main="Absolute correlation with QC features")
legend("topright", colnames(df)[3:6], fill=col1, border=col1, cex=1)

boxplot(zinb@W[,1]~batch, col=colPal1, ylab="PC1", xlab="Batch")
boxplot(zinb@W[,2]~batch, col=colPal1, ylab="PC2", xlab="Batch")

boxplot(zinb@W[,1]~position, col=colPal2, ylab="PC1", xlab="Position")
boxplot(zinb@W[,2]~position, col=colPal2, ylab="PC2", xlab="Position")
```

## Add batches in ZINB model

```{r zinb_batch}
zinb_batch <- zinbFit(raw, ncores = ncores, K = 2, X=model.matrix(~batch))
plot(zinb_batch@W, pch=19, col=colMerged, main="ZINB")
```

```{r zinb_batch_cov}
plot(zinb_batch@W, col=colBatch, pch=20, main="Colored by batch")

plot(zinb_batch@W, col=colPos, pch=20, main="Colored by position")

df <- data.frame(W1=zinb_batch@W[,1], W2=zinb_batch@W[,2], nmolecule=nmolecule, diameter=diameter, coverage=coverage, detection_rate=detection_rate) 

df %>% ggplot(aes(W1, W2)) + geom_point(aes(color=nmolecule)) + scale_colour_gradient(low="blue", high="yellow") + theme_classic()

df %>% ggplot(aes(W1, W2)) + geom_point(aes(color=diameter)) + scale_colour_gradient(low="blue", high="yellow") + theme_classic()

df %>% ggplot(aes(W1, W2)) + geom_point(aes(color=coverage)) + scale_colour_gradient(low="blue", high="yellow") + theme_classic()

df %>% ggplot(aes(W1, W2)) + geom_point(aes(color=detection_rate)) + scale_colour_gradient(low="blue", high="yellow") + theme_classic()

pairs(df, col=colMerged, pch=19)

cors <- sapply(1:2, function(i) cor(zinb_batch@W[,i], df[,3:6]))
colnames(cors) <- paste("PC", 1:NCOL(cors), sep="")
rownames(cors) <- paste("QC", 1:NROW(cors), sep="")
barplot(abs(cors), beside = TRUE, col=col1[1:4], border=col1[1:4], ylim=c(0, 1), space=c(0, 2), main="Absolute correlation with QC features")
legend("topright", colnames(df)[3:6], fill=col1, border=col1, cex=1)

boxplot(zinb_batch@W[,1]~batch, col=colPal1, ylab="PC1", xlab="Batch")
boxplot(zinb_batch@W[,2]~batch, col=colPal1, ylab="PC2", xlab="Batch")

boxplot(zinb_batch@W[,1]~position, col=colPal2, ylab="PC1", xlab="Position")
boxplot(zinb_batch@W[,2]~position, col=colPal2, ylab="PC2", xlab="Position")
```


# Gene-level covariates

```{r gc_content, eval=FALSE}
counts <- counts[bm[,1],]
gene_info = getGeneLengthAndGCContent(as.character(bm[,2]), "mm10", mode="org.db")
rownames(gene_info) = bm[,1]
gene_info = na.omit(gene_info)
```

```{r write_out}
save(zinb, zinb_batch, zifa_fq, zifa_tmm, zifa_tc, zifa_raw, pc_tmm, pc_fq, pc_tc, pc_raw, file="zeisel_covariates.rda")
```
